\textbf{Predicting virus-receptor mutant binding by molecular dynamics simulation}

### 2013-07-01 paper.tex
Updated paper and added mutant comparison boxplot.

\documentclass[12pt]{article}

@@

Existing computational methods to predict protein--protein interaction affinity often perform poorly in important test cases. In particular, the effects of multiple mutations, non-alanine substitutions, and flexible loops are difficult to predict with available tools and protocols. We present here a new method to interrogate affinity differences resulting from mutations in a host-virus protein--protein interface. Our method is based on extensive non-equilibrium all-atom simulations: We computationally pull the machupo virus (MACV) spike glycoprotein (GP1) away from the human transferrin receptor (hTfR1) and estimate affinity using the maximum applied force during a pulling simulation and the area under the force-versus-distance curve. We find that these quantities provide novel biophysical insight into the GP1/hTfR1 interaction. First, with no prior knowledge of the system we can differentiate among wild-type and mutant complexes. Second, although the static co-crystal structure shows two large hydrogen-bonding networks in the GP1/hTfR1 interface, our simulations indicate that only one of them is critical for the binding interaction. Third, one viral site known to be critical for infection may mark an important evolutionary suppressor site for infection-resistant hTfR1 mutants. Finally, our method provides an elegant framework to compare the effects of multiple mutations, individually and jointly, on protein--protein interactions.
\end{abstract}

@@

Here we propose a new method to probe mutational effects in protein--protein interactions directly, by pulling two proteins apart that are initially bound in a complex. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC).

@@

We applied our computational method to wild type (WT) and mutant complexes, and found that we could resolve relative differences in unbinding and predict significant affinity changes. At sites known to be important for successful viral entry, we found that the biochemical cause of reduced infectivity is often not as simple as the static structure suggests. For example, the static structure shows a hydrogen-bonding network connected to site Asn348 in hTfR1. According to our simulations, this network likely does not affect binding affinity directly. Our study offers an all-atom steered molecular dynamic approach to avoid the pitfalls of several existing methods used to evaluate mutations in protein--protein interfaces.

@@

Last, we further analyzed a single mutation in GP1, vArg111Ala. As mentioned in the previous subsection, in our simulations this mutant showed no significant change in either maximum applied force or AUC (Tables~\ref{tab:MAF_p} and~\ref{tab:AUC_p}), even though both quantities were, on average, lower than in WT (Table~\ref{tab:summary_statistics}). This result was somewhat surprising, since Tyr211Ala, presumably disrupting the same hydrogen-bonding network as vArg111Ala, displayed a significant reduction in affinity. To probe the interaction between position 111 in the GP1 and position 211 in the hTfR1 further, we also tested the double mutant vArg111Ala/Tyr211Ala. This double mutant showed affinity indistinguishable from WT and significantly higher than Tyr211Ala alone  (Table~\ref{tab:pairwise_differences}). Thus, vArg111Ala in the GP1 appeared capable of suppressing the loss-of-binding caused by Tyr211Ala. This result shows that the two sites do indeed interact significantly, and that replacing the hydrogen-bonding network at these sites with a hydrophobic interaction leads to comparable binding affinity.

@@

Our findings rationalize several effects observed in both infectivity data and rodent populations \citep{Rad2008,Rad20111}. First, we found that some substitutions at positions 211 and 348 did negatively affect the strength of receptor binding. However, the computational data suggest that the reason and nature of the effects at these two sites are very different. Congruent with entry data \citep{Rad2008,Rad20111}, mutations at Tyr211 often have a dramatic effect on receptor binding. Second, no individual point mutation at Asn348 showed a significantly different affinity from the WT complex. However, when Tyr211Ala was paired with large (Trp) and positively charged (Lys) substitutions at position 348, the result was a larger than expected synergistic difference. That is, the double mutant Tyr211Ala/Asn348Trp caused a much larger decrease in binding than we expected from either mutation individually. Third, the GP1 mutation vArg111Ala causes a loss-of-infection during \textit{in vitro} infectivity assays \citep{Rad20112}, yet it was indistinguishable from the WT complex in our simulations. Although Tyr211Ala was the most disruptive single mutant we tested, vArg111Ala in the GP1 was able to restore mean maximum applied force to WT levels (Table~\ref{tab:summary_statistics}), and to levels significantly higher than observed for Tyr211Ala alone. As mutations at hTfR1 site 211 are known to reduce infectivity \emph{in vitro} and \emph{in vivo}, it is possible that vArg111 in GP1 can serve as a site for compensatory MACV mutations.

%Structural analysis reveals that the biochemical cause involves the destruction of a critical hydrogen-bonding network with Ser113 and vArg111 in GP1. Further, Tyr211Asp and Tyr211Thr were easily resolved from the WT complex, but they appear to cause an increase in affinity. As a result, we predict experimental structural analysis will show that Tyr211Asp and Tyr211Thr mutations cause a relatively dramatic tertiary structure deviation. Given that our model starts from an associated complex, it is not surprising that Tyr211Asp would cause increased rather than decreased affinity. Aspartate is charged rather than polar at physiological pH and its shorter side chain means the proteins would be pulled in closer apposition. Nevertheless, our results show Tyr211, or some other amino acid that is capable of hydrogen bonding, is necessary for a stable hydrogen-bonding network leading to efficient binding. Moreover, the cause of decreased infectivity seen in rodent populations with the Tyr211Asp mutation is not a simple change in binding affinity.

%Previous work has shown that the hTfR1 mutation Asn348Lys is capable of producing decreased infectivity \emph{in vitro} \citep{Rad2008,Rad20111}. Moreover, the molecular model based on x-ray diffraction data suggests destruction of a second hydrogen-bonding network as the cause \citep{Abraham2010}. In our simulations no individual point mutation at Asn348 showed a significantly different affinity from the WT complex. Moreover, the mutation Asn348Ala resulted in an average maximum applied force that was virtually identical to WT hTfR1 (Table~\ref{tab:pairwise_differences}). Even when paired with the highly destructive Tyr211Ala mutation, Asn348Ala was refractory to changes in binding affinity. By contrast, when Tyr211Ala was paired with large (Trp) and positively charged (Lys) substitutions at position 348, the result was a larger than expected synergistic difference. That is, the double mutant Tyr211Ala/Asn348Trp caused a much larger decrease in binding than we expected from either mutation individually. Moreover, although Asn348Trp and Asn348Lys individually failed to be distinguishable from WT hTfR1, they both caused a mean reduction in maximum applied force and produced larger than expected synergistic effects. This data suggests that site 348 is size and charge restricted. Any mutation to a very large or positively charged side chain will likely produce decreased infectivity.

@@

%Making accurate predictions regarding the effects of non-alanine mutations at protein--protein interfaces remains  challenging. The main obstacle is the enormous number of chemically allowed three dimensional configurations in proteins \citep{Gumbart2012}. In theory, one could carry out all-atom simulations from physical first principles \citep{Gumbart2012} for very long time periods. After equilibrium is reached in such simulations, one could then compute a binding constant by observing the fractional association/dissociation time among two proteins. Unfortunately, due to limited computational power and available methods, reaching equilibrium with a simulated protein system is truly a herculean task and is essentially impossible at this time \citep{Gumbart2012}. There are two broadly useful computational approaches that avoid first-principles equilibrium simulations. First, to reduce computational complexity several studies have used rigid body docking and machine learning \citep{Vreven2011,Vreven2012,Bajaj2011,Hwang2010}. While these approaches can work well in certain scenarios, they often underperform in applications to novel systems, due to their dependence on available training data. Second, hybrid approaches exist that exploit non-equilibrium calculations within the context of all atom simulations \citep{Dixit2001,Is2001A,Is2001B,Park2004,Grant2011,Kortemme2004,Gumbart2012}. However, to date most of these methods have not been applied to protein--protein interactions, but rather to simpler scenarios such as protein--ligand interactions. What we have shown here is that hybrid non-equilibrium approaches can provide a useful and practical tool in studying protein--protein interactions.

We have probed the biochemical effects of mutations at the host-virus interface between MACV and its human receptor. We applied an approach that imparts forces in an all-atom molecular dynamics (MD) simulation to drive the system through equilibrium. Our work has shown that with relatively modest computing resources it is possible to use MD simulations to investigate point mutations in the GP1/hTfR1 interface. In most cases, a double mutant was easily sufficient to resolve a difference in relative affinity. Moreover, relative to the WT complex one point mutant was capable of producing a large decrease in binding affinity. We applied our technique to the WT complex and 10 mutations in the GP1/hTfR1 interface. We were able to rationalize the results of infectivity data with fundamental physical simulations. Our data has provided some direction for further biochemical investigation in this system. A polar residue is absolutely critical at position 211 in hTfR1. Any mutation to a nonpolar residue will probably lead to a rapid viral compensatory response. Also, although rodent populations may produce further mutations at Asn348 in hTfR1, unless a mutation is positively charged or very large it is unlikely to result in a large disruption to receptor binding. We have also provided both a site (vArg111) and a mutation (vArg111Ala) that may be critical for future host-range adaptation of MACV, and thus should be monitored in surveillance efforts. Finally, our analysis offers a new means to systemically investigate host-virus interactions through computational molecular dynamics.

>>>>

\documentclass[12pt]{article}
\pdfoutput=1

@@

Existing computational methods to predict protein--protein interaction affinity often perform poorly in important test cases. In particular, the effects of multiple mutations, non-alanine substitutions, and flexible loops are difficult to predict with available tools and protocols. We present here a new method to interrogate affinity differences resulting from mutations in a host-virus protein--protein interface. Our method is based on extensive non-equilibrium all-atom simulations: We computationally pull the machupo virus (MACV) spike glycoprotein (GP1) away from the human transferrin receptor (hTfR1) and estimate affinity using the maximum applied force during a pulling simulation and the area under the force-versus-distance curve. We find that these quantities can provide novel biophysical insight into the GP1/hTfR1 interaction. First, with no prior knowledge of the system we can differentiate among wild-type and mutant complexes. Second, although the static co-crystal structure shows two large hydrogen-bonding networks in the GP1/hTfR1 interface, our simulations indicate that only one of them is critical for the binding interaction. Third, one viral site known to be critical for infection may mark an important evolutionary suppressor site for infection-resistant hTfR1 mutants. Finally, our method provides an elegant framework to compare the effects of multiple mutations, individually and jointly, on protein--protein interactions.
\end{abstract}

@@

Here we propose a new method to probe mutational effects in protein--protein interactions directly based on first principles simulation, by pulling two proteins apart that are initially bound in a complex. We impart a pulling force within an all-atom molecular dynamics simulation on one member of the complex while the other is held in place. Then, we measure the force required for dissociation. As a proxy for relative binding affinity, we use the maximum applied force required for separation and the area under the force-versus-distance curve (AUC).

@@

We applied our computational method to wild type (WT) and mutant complexes, and found that we could resolve relative differences in unbinding and predict significant affinity changes. At sites known to be important for successful viral entry, we found that the biochemical cause of reduced infectivity may not be as simple as the static structure suggests. For example, the static structure shows a hydrogen-bonding network connected to site Asn348 in hTfR1. According to our simulations, this network likely does not affect binding affinity directly. Our study offers an all-atom steered molecular dynamic approach to avoid the pitfalls of several existing methods used to evaluate mutations in protein--protein interfaces.

@@

Last, we further analyzed a single mutation in GP1, vArg111Ala. As mentioned in the previous subsection, in our simulations this mutant showed no significant change in either maximum applied force or AUC (Tables~\ref{tab:MAF_p} and~\ref{tab:AUC_p}), even though both quantities were, on average, lower than in WT (Table~\ref{tab:summary_statistics}). This result was somewhat surprising, since Tyr211Ala, presumably disrupting the same hydrogen-bonding network as vArg111Ala, displayed a significant reduction in affinity. To probe the interaction between position 111 in the GP1 and position 211 in the hTfR1 further, we also tested the double mutant vArg111Ala/Tyr211Ala. This double mutant showed affinity indistinguishable from WT and significantly higher than Tyr211Ala alone  (Table~\ref{tab:pairwise_differences}). Thus, vArg111Ala in the GP1 appeared capable of suppressing the loss-of-binding caused by Tyr211Ala. This result shows that the two sites do indeed interact significantly, and that replacing the hydrogen-bonding network at these sites with a hydrophobic interaction could lead to comparable binding affinity.

@@

Our findings may rationalize several effects observed in both infectivity data and rodent populations \citep{Rad2008,Rad20111}. First, we found that some substitutions at positions 211 and 348 did affect the strength of receptor binding. However, the computational data suggest that the reason and nature of the effects at these two sites are very different. Congruent with entry data \citep{Rad2008,Rad20111}, mutations at Tyr211 often have a dramatic effect on receptor binding. Second, no individual point mutation at Asn348 showed a significantly different affinity from the WT complex. However, when Tyr211Ala was paired with large (Trp) and positively charged (Lys) substitutions at position 348, the result was a larger than expected synergistic difference. That is, the double mutant Tyr211Ala/Asn348Trp caused a much larger decrease in binding than we expected from either mutation individually. Third, the GP1 mutation vArg111Ala causes a loss-of-infection during \textit{in vitro} infectivity assays \citep{Rad20112}, yet it was indistinguishable from the WT complex in our simulations. Although Tyr211Ala was the most disruptive single mutant we tested, vArg111Ala in the GP1 was able to restore mean maximum applied force to WT levels (Table~\ref{tab:summary_statistics}), and to levels significantly higher than observed for Tyr211Ala alone. As mutations at hTfR1 site 211 are known to reduce infectivity \emph{in vitro} and \emph{in vivo}, it is possible that vArg111 in GP1 can serve as a site for compensatory MACV mutations.

@@

We have probed the biochemical effects of mutations at the host-virus interface between MACV and its human receptor. We applied an approach that imparts forces in an all-atom molecular dynamics (MD) simulation to drive the system through equilibrium. Our work has shown that with relatively modest computing resources it is possible to use MD simulations to investigate point mutations in the GP1/hTfR1 interface. In most cases, a double mutant was easily sufficient to resolve a difference in relative affinity. Moreover, relative to the WT complex one point mutant was capable of producing a large decrease in binding affinity. We applied our technique to the WT complex and 10 mutations in the GP1/hTfR1 interface. Our data has provided some direction for further biochemical investigation in this system. A polar residue is absolutely critical at position 211 in hTfR1. Any mutation to a nonpolar residue will probably lead to a rapid viral compensatory response. Also, although rodent populations may produce further mutations at Asn348 in hTfR1, unless a mutation is positively charged or very large it is unlikely to result in a large disruption to receptor binding. Finally, our analysis offers a new means to systemically investigate host-virus interactions through computational molecular dynamics.
